TODO: title

### 2014-05-03 paper/main.tex
some work on outlineing and small eval changes

\usepackage{hyperref}
%  \multicolumn{2}{|c|}{Team sheet} \\

@@

  % new user experience is important, and here is why.
    % talk about first impressions and user trust
  How a recommender system handels new users can change how these users interact with the recommender system.

%%%%%%%%%%%%%%%%%
  The new user experience is a very important part of any recommender system.
  The first interaction with a system is the only chance the system gets to make a first impression on the user.
  This first impression can be very important in structuring a user's expectations for how the system will behave, and determining how much the user trusts the system's recommendations.
  If the system behaves poorly when the user is new, then the user may leave and not come back.
  This is especially a concern with recommender systems which rely on large amounts of user information to make personalization decisions.
  When users are new the system will not have much information to personalize the user experience which may lead the user to leave the system.
%%%%%%%%%%%%%%%%%%%%%%

  % Past work on cold start has focused on… (how long should this section be?)
    % ways to sort items to get to most value out of early ratings
    % research looking at user experience and other forms of data collection

@@

  % One common new user strategy is to have a special data collection mode.
  % collect ratings fast to get over a precived ``bad period''
  % work on this focuses on how to collect useful ratings.
  % (framework from decision tree work)
  The best studied strategy for new users is to have users go through a new user survey.

  % One promising approach is to let user control interaction
  % user or system driven control.
  % promising, we don't know much about follow up.
  One promissing approach to new users is to have users control the new user survey.

  % How an algorithm behaves the first time a user sees it is important (first impression)
    % Past work looks carefully at issues X, Y, and Z, but doesn’t take a systematic look at the behavior of algorithms for users with few ratings.
  % Our work sits at the intersection of two important fields of recommender system research: New user experiance, and comparing algorithms. While there has been much work comparing various algorithms, there has been little formal work comparing algorithms based on their performance for new users. Therefore our work focuses on the behavior of recommender algorithms across different amounts of user data. (subsection might need re-ordering)

  Our work sits at the intersection of two important traditions of recommender systems research: new user experience, and offline algorithm comparison.
  Our goal is to understand how different algorithms behave for new users, especially how their behavior changes as new users add ratings into the system.
  While there has been much work comparing algorithms in an offline setting, very little of it has explored what are the best algorithms for users with very few ratings.
  Furthermore, we do not seek to find the algorithm that scores best on some metric of choice, instead, we seek to form a complete understanding of the differences between algorithms for cold start use.

  We structure our work around the following research questions.
  \begin{itemize}
    \item RQ1: How do algorithms behave for users with very few ratings?
    \item RQ2: How does algorithm behavior change with increasing ratings for new users?
  \end{itemize}
  Unfortunately, algorithm behavior is a broad and hard to define subject.
  We will look at three distinct ways to quantify the behavior of a recommender system, each of which may lead to multiple metrics.
%  One of the most traditional ways to measure the behavior of a recommendation system is to measure its performance on the rating task.
%  This is normally done with metrics such as RMSE or MAE.
%  The accuracy of the algorithm may be important at cold start as users my judge the recommender on how well it predicts items they have not yet rated.
%  This early judgment may be very important to the user.

%  Recently, work in the recommender system community has been shifting away from prediction accuracy to start looking at how well the system ranks and recommends items.
%  This change was, in part prompted by work criticizing the practice of measuring only the RMSE of recommender systems \cite{TODO:rmse considered harmful...}
%  The quality of the ranking provided by the recommender has normally been measured with information retrial metrics such as nDCG.
%  The quality of the recommendations has normally been measured by looking at which sort of items end up being recommended with metrics such as precision@N, recall@N, fallout@N etc.
%  We will also look at some simpler metrics along with these such as simply evaluating the ratings and predictions that end up in top 20 recommendations.

%  Finally, we find that

@@

  Prediction accuracy, as measured by RMSE or MAE is one of the classic techniques in recommended system evaluation.
  The idea behind these evaluations is to measure the accuracy of predictions on withheld recommendations.
  If the algorithm can successfully predict the withheld ratings, then we expect it will accurately predict future ratings as well.
  Because users may build trust in the system by looking at the predictions of items, we are interested in knowing which algorithms are most accurate, and which show the most notable change in accuracy as ratings are entered.

  More recently, work as shifted away from accuracy metrics in favor of metrics that evaluate a recommender systems ability to rank, or recommend.
  % TODO: cite johnjoe here?
  These algorithms try to evaluate how useful the recommendations provided by the system are to the user.
  This is commonly done by looking either at how withheld items are ranked with metrics such as nDCG, or by looking at if good or bad withheld items tend to be recommended positively by the recommender.

  Prior work \cite{ziegler and martijn} has shown that users are sensitive to properties of recommendations that are not as trivially good or bad as those above.
  We will look at three properties that can be measured in an offline context, the popularity, diversity, and spread of recommendations.

  We expect that different algorithms will recommend items that are more or less popular.
  This difference could explain feelings that a list isn't personal enough (too popular), or is too novel / obscure (not popular enough).

  Past work \cite{ziegler and martijn} has shown that users tend to prefer recommendation lists that are more diverse.
  <TODO: skim this past work again and report some details/results.>

  Finally, we expect that algorithms with a good understand of its users will be able to recommend different items to different users.
  To measure this we will look at how well the recommender spreads its attention of many items.
  We will do this by computing the Shannon's entropy of the items recommended to each user.
  If the recommender system recommends the same items to each user, then the entropy will be low.
  Conversely, if the recommender recommends different items to each user, then the entropy will high.
  In practice, a recommender may focus on both a shared core of popular or well reviewed items, and more personal items, which would lead to middle values.

  Based on this list of possible metrics, we will clarify our research questions with three sub questions:
  \begin{\itemize}
  \item How do algorithms behave with respect to prediction accuracy?
  \item How do algorithms behave with respect to ranking items appropriately?
  \item How do algorithms behave as measured by other metrics such as popularity and diversity?
  \end{\itemize}

@@

  To answer our research questions we will evaluate algorithms in an offline simulation designed to simulate users as they join the system.
  We will do this by running separate instances of the standard train and test methodology, each one designed so that test users have some small number of ratings.
  We will then plot metric results against the number of retained items to understand both how the algorithms compare on these metrics, and how the algorithm performance changes with the number of ratings provided.
  This, or a similar technique, has been used by previous authors \cite{DrennerInitialExperiance, TenIsEnough, AdaptiveBootstrap}, however, these normally are used with only one algorithm or metric, and normally just to make the point that prediction accuracy improves as the number of ratings grows.
  We will use this technique to develop a much more nuanced understanding of the differences between algorithms, and how they behave for users with few ratings.

@@

  We seek to develop an understanding of how a range of different standard algorithms perform at for new users.
  Rather than picking advanced algorithms that are tuned for new user performance, we decided to first understand how representatives of three common types of algorithms perform.
  Therefore we will compare User-User, Item-Item, and Funk's SVD algorithms against two simple baseline algorithms.
  More information about these algorithms can be found in table \ref{tbl:algo}
  Code for our evaluation, using the Lenskit evaluation framework \cite{lenskit} is available at \url{https://bitbucket.org/kluver/coldstartrecommendation}.

@@

  Each algorithm was tuned using the values from Ekstrand's 2012 lenskit evaluation \cite{ekstrand2012recommenders}.
  Sine this evaluation was on a slightly different dataset (MovieLens 10M), we performed only minor tuning from these parameters.
  Most notably we found the tuning of the damping parameter used in the baseline recommenders to be too large, which limited accuracy at cold start.
  Overall, we did not find our qualitative results to be highly sensitive to minor tuning changes, although specific values may change slightly.

  \subsection*{Metrics}

@@

  % dataset
    % # user
    % # items
    % distributions of # ratings per user (important: min 20)

  % temporal replay evaluation 
    % similar technique, harder to do than ours, and with a different goal.

>>>>

\usepackage{hyperref}

@@

  % new user experience is important, and here is why.
    % talk about first impressions and user trust
    % How an algorithm behaves the first time a user sees it is important (first impression)
  How a recommender system handels new users can change how these users interact with the recommender system.
%%%%%%%%%%%%%%%%%%%%%%
  The new user experience is a very important part of any recommender system.
  The first interaction with a system is the only chance the system gets to make a first impression on the user.
  This first impression can be very important in structuring a user's expectations for how the system will behave, and determining how much the user trusts the system's recommendations.
  If the system behaves poorly when the user is new, then the user may leave and not come back.
  This is especially a concern with recommender systems which rely on large amounts of user information to make personalization decisions.
  When users are new the system will not have much information to personalize the user experience which may lead the user to leave the system.
%%%%%%%%%%%%%%%%%%%%%%

@@

  % One common new user strategy is to have a special data collection mode.
  % collect ratings fast to get over a precived ``bad period''
  % work on this focuses on how to collect useful ratings.
  % (framework from decision tree work)
  The best studied strategy for new users is to have users go through a new user rating survey.

  % One promising approach is to let user control interaction
  % user or system driven control.
  % promising, we don't know much about follow up.
  One promissing approach to new users is to have users control the new user rating survey.

  % bridge from past work to our work.
    % point out holes (evaluating standard algorithms for little to no ratings, wholeistic evaluation)
    % point out that other approaches than smart object ordering exist, and we need to be able to compare.
    % cite the comparison work at CHI.
    % we propose to evaluate 
  While past work has covered the new user rating survey as a solution to new user cold start, much less work has been done comparing other strategies.

  % Past work looks carefully at issues X, Y, and Z, but doesn’t take a systematic look at the behavior of algorithms for users with few ratings.
  % Our work sits at the intersection of two important fields of recommender system research: New user experiance, and comparing algorithms. While there has been much work comparing various algorithms, there has been little formal work comparing algorithms based on their performance for new users. Therefore our work focuses on the behavior of recommender algorithms across different amounts of user data. (subsection might need re-ordering)

  Our work sits at the intersection of two important traditions of recommender systems research: new user experience, and offline algorithm comparison.
%%%%%%%%%%%%%%%%%%%%%%
  Our goal is to understand how different algorithms behave for new users, especially how their behavior changes as new users add ratings into the system.
  While there has been much work comparing algorithms in an offline setting, very little of it has explored what are the best algorithms for users with very few ratings.
  Furthermore, we do not seek to find the algorithm that scores best on some metric of choice, instead, we seek to form a complete understanding of the differences between algorithms for cold start use.
%%%%%%%%%%%%%%%%%%%%%%

  We structure our work around the following research questions.
  \begin{itemize}
    \item RQ1: How do algorithms behave for users with very few ratings?
    \item RQ2: How does algorithm behavior change with increasing ratings for new users?
  \end{itemize}
  Unfortunately, algorithm behavior is a broad and hard to define subject.
  We will look at three distinct ways to quantify the behavior of a recommender system, each of which may lead to multiple metrics.

@@

  Prediction accuracy, as measured by RMSE or MAE is one of the classic techniques in recommended system evaluation.
%%%%%%%%%%%%%%%%%%%%%%
  The idea behind these evaluations is to measure the accuracy of predictions on withheld recommendations.
  If the algorithm can successfully predict the withheld ratings, then we expect it will accurately predict future ratings as well.
  Because users may build trust in the system by looking at the predictions of items, we are interested in knowing which algorithms are most accurate, and which show the most notable change in accuracy as ratings are entered.
%%%%%%%%%%%%%%%%%%%%%%

  More recently, the research community has shifted away from accuracy metrics in favor of metrics that evaluate a recommender system's ability to rank or recommend.
%%%%%%%%%%%%%%%%%%%%%%
  % TODO: cite johnjoe here?
  These algorithms try to evaluate how useful the recommendations provided by the system are to the user.
  This is commonly done by looking either at how withheld items are ranked with metrics such as nDCG, or by looking at if good or bad withheld items tend to be recommended positively by the recommender.
%%%%%%%%%%%%%%%%%%%%%%

  Prior work has shown that users are sensitive to properties of a recommender other than simply how well it performs.
%%%%%%%%%%%%%%%%%%%%%%
  These properties, such as the diversity of the output recommendations, can effect how the user interacts with the recommender and drastically change the user experience \cite{ziegler and martijn}.
  Unfortunately, there is no way to know all such interesting metrics
  We will look at three such metrics, the popularity, diversity, and spread of the recommendations.
%%%%%%%%%%%%%%%%%%%%%%

  The popularity of the items recommended to a user can have an impact on the user's satisfaction.
%%%%%%%%%%%%%%%%%%%%%%
  If the user is only recommended items that are very popular, then the recommender is likely failing to make novel and interesting recommendations.
  In this case the user may feel that the recommender is not helpful in finding new and interesting things to watch.
  However, if a new user is only recommended items that are not well known, it is unlikely that the user will have heard of the items recommended.
  This could cause the user to not trust the system, as they have no way of comparing the recommendations against their personal knowledge.
%%%%%%%%%%%%%%%%%%%%%%

  Past work \cite{ziegler and martijn} has shown that users tend to prefer recommendation lists that are more diverse.
%%%%%%%%%%%%%%%%%%%%%%
  <TODO: skim this past work again and report some details/results.>
%%%%%%%%%%%%%%%%%%%%%%

  Finally, we expect that algorithms with a good understand of its users will be able to recommend different items to different users.
%%%%%%%%%%%%%%%%%%%%%%
  To measure this we will look at how well the recommender spreads its attention of many items.
  We will do this by computing the Shannon's entropy of the items recommended to each user.
  If the recommender system recommends the same items to each user, then the entropy will be low.
  Conversely, if the recommender recommends different items to each user, then the entropy will high.
  In practice, a recommender may focus on both a shared core of popular or well reviewed items, and more personal items, which would lead to middle values.
%%%%%%%%%%%%%%%%%%%%%%

  Based on this list of possible metrics, we will clarify our research questions with three sub questions:
  \begin{itemize}
  \item How do algorithms behave with respect to prediction accuracy?
  \item How do algorithms behave with respect to ranking items appropriately?
  \item How do algorithms behave as measured by other metrics such as popularity and diversity?
  \end{itemize}

@@

  To answer our research questions we will evaluate algorithms in an offline simulation designed to simulate users as they join the system.
%%%%%%%%%%%%%%%%%%%%%%
  We will do this by running separate instances of the standard train and test methodology, each one designed so that test users have some small number of ratings.
  We will then plot metric results against the number of retained items to understand both how the algorithms compare on these metrics, and how the algorithm performance changes with the number of ratings provided.
  This, or a similar technique, has been used by previous authors \cite{DrennerInitialExperiance, TenIsEnough, AdaptiveBootstrap}, however, these normally are used with only one algorithm or metric, and normally just to make the point that prediction accuracy improves as the number of ratings grows.
  We will use this technique to develop a much more nuanced understanding of the differences between algorithms, and how they behave for users with few ratings.
%%%%%%%%%%%%%%%%%%%%%%

@@

  We seek to develop an understanding of how a range of different standard algorithms perform at for new users.
%%%%%%%%%%%%%%%%%%%%%%
  Rather than picking advanced algorithms that are tuned for new user performance, we decided to first understand how representatives of three common types of algorithms perform.
  Therefore we will compare User-User, Item-Item, and Funk's SVD algorithms against two simple baseline algorithms.
  More information about these algorithms can be found in table \ref{tbl:algo}
  Code for our evaluation, using the Lenskit evaluation framework \cite{lenskit} is available at \url{https://bitbucket.org/kluver/coldstartrecommendation}.
%%%%%%%%%%%%%%%%%%%%%%

@@

  Each algorithm was tuned using the values from Ekstrand's 2012 lenskit evaluation \cite{ekstrand2012recommenders}.
%%%%%%%%%%%%%%%%%%%%%%
  Since this evaluation was on a slightly different dataset (MovieLens 10M), we performed only minor tuning from these parameters.
  Most notably we found the tuning of the damping parameter used in the baseline recommenders to be too large for this dataset.
  As the damping can have a large effect on baseline quality for new users, we adjusted to a more reasonable damping for this dataset.
  Overall, we did not find our results to be highly sensitive to minor tuning changes, although specific values may change slightly.
%%%%%%%%%%%%%%%%%%%%%%

  \subsection*{Metrics}
  We will discuss a total of eleven metrics.
%%%%%%%%%%%%%%%%%%%%%%
  Some of these metrics will prove to be redundant, and future work should be able to use a much smaller list.
  The metrics are listed in table \ref{tbl:metrics}.
  \begin{table}[ht!]
    \centering
    \begin{tabular}{|p{7em}|p{16em}|}
      \hline
      \multicolumn{2}{|c|}{{\bf Accuracy Metrics}} \\\hline
      RMSE                   & Standard prediction accuracy metric. \\\hline
      \hline
      \multicolumn{2}{|c|}{{\bf Recommendation Metrics}} \\\hline
      nDCG                   & Standard ranking quality evaluation.\\\hline
      Precision@20           & Count of recommended items in the test set and rated 4.0 or higher.\\\hline
      MAP@20                 & Average of precision at each rank one through twenty.\\\hline
      Fallout@20             & Count of recommended items in the test set and rated 2.0 or lower.\\\hline
      SeenItems@20           & Count of recommended items in the test set.\\\hline
      MeanRating@20          & Average rating of recommended items that are in the test set.\\\hline
      RMSE@20                & RMSE computed only over those items that were recommended.\\\hline
      \hline
      \multicolumn{2}{|c|}{{\bf Other Metrics}} \\\hline
      Average\-Popularity@20 & The average number of users in the training set that have seen recommended items.\\\hline
      AILS@20                & The average pairwise similarity between recommended items. \\\hline
      Spread@20              & The entropy of the distribution of recommended items for users in the test set.\\\hline
    \end{tabular}
    \caption{TODO:}
    \label{tbl:metrics}
  \end{table}

  All metrics evaluating recommendations (all but RMSE and nDCG) are based on the top 20 predicted items for the user, excluding items in the training set.
  The AILS@20 metric is based on the ILS metric introduced by ziegler et al \cite{ziegler}.
  AILS@N is simply a rescaling of ILS to be scale free with regards to N, this was done to make the metric easier to interpret, by putting it on the same scale as similarity values.
  The choice of 20 items for the evaluations was arbitrary, testing with other values showed similar results, except on one metric.
  This issue will be addressed later in the results section.

%This is simply a rescaling of the ILS metric from \cite{ziegler}. Similarity is computed by subtracting the average item rating from the item vectors for each item and then computing the cosine similarity.\\\hline

@@

  \subsection*{Dataset}
  % dataset
    % # user
    % # items
    % distributions of # ratings per user (important: min 20)

  \subsection*{Evaluation Procedure}
